Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-16 response.tex

We have modified the introduction in order to define our goals more
clearly.

\emph{I think that the possibility to estimate the best prediction
  error in advance from a relatively simple WPE characteristic would
  be interesting and useful, if it could hold true for a wide class of
  systems. Such a question would deserve a careful study as that
  started in the presented work. However, I have a number of critical
  remarks and strongly doubt that the conjecture of the authors can be
  valid for a reasonably wide class of systems (not even speaking of
  ``all systems''). Other aspects of the paper are not so important,
  in my view, despite some of them are fresh and interesting
  (e.g. taking a computer as an object of prediction). My critical
  remarks are listed below.}

Computer dynamics spans the full range of behavior reported in the
nonlinear dynamics literature.  Standard nonlinear time-series
analysis of data from these systems has confirmed the existence of
fixed points, periodic orbits of different periods, chaotic
trajectories with various Lyapunov exponents, etc.  in their dynamics.
While it is certainly possible that the nonlinear dynamics of computer
performance is subtly different from the nonlinear dynamics of other
systems---in a manner that cannot be detected by traditional nonlinear
time-series analysis---we believe that our results will hold for time
series drawn from other physical systems.

We are certainly aware that this claim may not be true, however, and
we are in the process of the follow-up study suggested by the
reviewer.  We have found that the relationship posited in our paper
holds for a time series of voltages from a chaotic laser used in the
SFI prediction competition, for instance.  (It also holds for various
paradigmatic systems, such as the Henon map and the Lorenz system, as
discussed in a later section of this response.)  The image below is an
augmented version of our revised Figure 7 that includes all of these
examples.

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
\end{center}

@@

It is certainly true that the naive and random-walk predictors can be
seen as subclasses of the ARIMA model family.  We have explicitly
included them, however, because they are used (sometimes to great
success) in practice and because that made for a more comprehensive
exploration.

The ``AutoARIMA'' method that was used to build all of the ARIMA
models in this paper---another ``off the shelf'' method in the
modeling and prediction communities---follows the procedure outlined
in Section \alert{[[blah]]} to choose values for the model parameters.
Sometimes the choices made by that procedure (e.g., setting a
particular parameter to zero) produce a naive or random-walk
predictor.  Our method can help practioners know whether this is a
good or bad thing.  If a particular model performs poorly (compared to
what the WPE value might suggest), it may be an indication that the
chosen parameter values were suboptimal---or even that no member of
the ARIMA family will work and one should try a nonlinear model.

% Unfortunately the WPE value alone is insufficient to differentiate
% these two situations.  It can identify both, but cannot distinguish.

We have changed the term used in the paper to {\tt auto.arima} in
order to make it clear that this refers not to the general class of
these models, but to a particular member of that class chosen via a
particular procedure.

@@

\smallskip

\noindent\emph{General remarks.}

\emph{I) Strict relationship between WPE and prediction error
  applicable to all systems is hardly possible. In particular, there
  is a notion of epsilon-entropy which tends to the KS entropy if
  epsilon (the cell size) tends to zero (see e.g.  [Boffetta et al.]
  cited above). For stochastic systems, the KS entropy is infinite and
  epsilon-entropy varies strongly with the scale epsilon. Seemingly,
  the permutation entropy estimate in this case will strongly depend
  on the choice of the word length L. The authors concentrate on L = 6
  and obtain coefficients of the regression between WPE and prediction
  error only for that choice and for their particular
  signals. However, other systems and other choices of L may easily
  lead to other regression coefficients so that deciding about the
  best possible prediction error from WPE seems impossible in
  general. It is good if such a decision is possible for a certain
  narrow class of systems. However, the authors do not even discuss to
  which class of systems their results apply.}

WPE values depend not only on wordlength, but also on \emph{data}
length---since one needs more data to get good statistics on the words
for the PE calculation.  These two interacting knobs are not easy to
set.  Perhaps as a result of this, wordlength choices are rarely
justified in the permutation entropy literature.

The extent of the discussion in [Bandt \& Pompe, {\sl PRL}, 2002], for
instance---the paper in which PE was proposed---is ``Our entropies are
calculated for different [$\ell$] but we do not to determine a limit
for large [$\ell$], although this is an interesting theoretical
problem.  ``For practical purposes, we recommend 3\dots7...''  The
corresponding discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads
``In their paper, Bandt \& Pompe recommended [$\ell$] of 3-7.  We
often found that 3 and 4 may still be too small, and a value of 5, 6,
or 7 seems to be the most suitable....''

@@

The right thing to do, as we argued in Section [[what]], is to compute
PE across a \emph{range} of wordlengths ($\ell$) and look for a flat
spot in the curve of PE versus $\ell$.  We used that method to choose
individual $\ell$ values for each time series studied here.  We
conjecture that a systematic approach like sill mitigate any variation
in the results, thereby preserving the shape of the WPE vs. MASE
curve.  We have not yet confirmed that conjecture, but we are in the
process of doing so.

Regarding generality: the goal of this paper was to explore model-free
heuristics for empirical time-series data, where one does not know
anything about the underlying system.  In situations like this,
equivalence classes are difficult to define.  We addressed
this---insofar as is possible in an experimental context---by choosing
a range of time-series data sets whose behavior spans a wide range of
dynamical behaviors.  For all of these examples---as well as for the
39 additional ones\footnote{Henon, Lorenz, SFI A, random walk,
  nonlinear transformations of the 30 [[svdtwo]] and [[svdsix]]
  signals} constructed for this rebuttal letter---the pattern in
Figure 7 remained virtually unchanged, as described in the following
entry.  This strongly suggests, but of course does not prove, that our
approach is general.  As the reviewer suggested, this will require
extensive study and evaluation in future work.

@@

\smallskip

\emph{II) PE is invariant under an invertible nonlinear change of
  variables. The prediction error measured as the authors suggest is
  not invariant. Thus, such a simple change of variables can lead, at
  least, to a change in regression coefficient between WPE and
  prediction error. Thus, the authors should also discuss to which
  variable for that class of systems their result apply. I think that
  the fact that the authors observe a certain regression line at all
  seems to be due to small number of different signals they consider
  (only 7 or 8 different data points, such a small number of points
  may lie approximately near a straight line by chance).}

\alert{Liz to fix following text, now that we have these experiments}

PE is invariant under an invertible nonlinear transformation of the
values in the time series, but WPE is not.  But MASE varies \emph{in a
  complementary way}.  Ryan will do $x(t)$ and $5e^{x(t)}$ and show
that it's on the curve.

% Ryan will insert figure here with x(t) = svdone and svdsix
% \begin{center}
%     \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
% \end{center}

This is consistent with our claim of generality; this is just another
time series.  A xform like this effectively changes the generating
process, but our model-free result handles it.

@@

The fit of the regression line was not a function of the small number
of points; when we included all 105 WPE-MASE pairs, rather than just a
representative from each program, the slope of the fit changed by 2\%
and the y intercept by 1\%.

We have revised the Figure, and the associated text, to include all
105 points.  We have also moved from a linear fit to a log fit in this
curve to better match the theoretical relationship between MASE and
WPE, as mentioned above.  Finally, we now show a ``fuzzy'' version of
the fitted curve in an effort to emphasize that the fundamental claim
of the paper is not related to the crispness of the fit.

\alert{Need the change and the confidence stuff for this: Fit changed
  by [[what]] when we redid it with these 34 new points.}  It now
reflects a map, a flow, a purely stochastic signal, a variety of
real-world data from two different systems (laser, computer), and
nonlinear transformations of that real-world data.  Seems pretty
compelling.

### 2014-07-16 intro.tex
fixing autoarima command and formatting bugs
formatting errors
revisions after Trident

\section{Introduction  }\label{sec:intro}
%****NOTES START HERE****

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a low-entropy
periodic signal is on the low end of the complexity spectrum.
\alert{This differs from the notion of complexity used by
  e.g. \cite{Shalizi2008}, which would consider a time series without
  any statistical regularities to be non-complex.  Liz moved this
  here.  Is there a better place for it?}  We argue that an extension
of \emph{permutation entropy}~\cite{bandt2002per}---a method for
approximating the entropy through ordinal analysis---is an effective
way to assess the complexity of a given time series.  Permutation
entropy is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  We chose to
study sensor data from a computer-performance experiment.  While this
is not a common laboratory experiment, it is a highly appropriate
choice here.  Computers are extremely complicated systems and their
dynamics is surprisingly rich.
% Modern microprocessor chips contain multiple processing units and
% multi-layer memories, for instance, and they use complicated
% hardware/software strategies to maximize performance by moving data
% and threads of computation across those resources.  As a result, the
The processor and memory loads during the execution of even a very
simple program can exhibit dynamical chaos, for
instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
example: a performance trace of a short program that repeatedly
initializes the upper triangle of a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a simple C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
A small change in the code can cause the dynamics to bifurcate to a
periodic regime.  
% 
% Indeed, many computer performance traces exhibit
% interesting regime changes as a program moves through the different
% phases of its operation.
By running different programs on the same computer, then, we can
produce traces that span the whole range of the complexity spectrum,
from completely predictable to completely unstructured---which makes
this an ideal testbed for this study\footnote{Predicting the
  \emph{state} of a computer, of course, would amount to solving the
  halting problem.  What we are doing here is predicting computer
  \emph{performance}, which does not violate the Rice-Shapiro
  theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

\section{Introduction}\label{sec:intro}
%****NOTES START HERE****

@@

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a low-entropy
periodic signal is on the low end of the complexity spectrum.  This
differs from the notion of complexity used by e.g. \cite{Shalizi2008},
which would consider a time series without any statistical
regularities to be non-complex.  We argue that an extension of
\emph{permutation entropy}~\cite{bandt2002per}---a method for
approximating the entropy through ordinal analysis---is an effective
way to assess the complexity of a given time series.  Permutation
entropy is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  We chose to
study sensor data from a computer-performance experiment.  While this
is not a common laboratory experiment, it is a highly appropriate
choice here.  Computers are extremely complicated systems and their
dynamics is surprisingly rich.
% Modern microprocessor chips contain multiple processing units and
% multi-layer memories, for instance, and they use complicated
% hardware/software strategies to maximize performance by moving data
% and threads of computation across those resources.  As a result, the
The processor and memory loads during the execution of even a very
simple program can exhibit dynamical chaos, for
instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
example: a performance trace of a short program that repeatedly
initializes the upper triangle of a matrix in column-major order.
%
 \begin{figure}[htp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a simple C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
A small change in the code can cause the dynamics to bifurcate to a
periodic regime.  
% 
% Indeed, many computer performance traces exhibit
% interesting regime changes as a program moves through the different
% phases of its operation.
By running different programs on the same computer, then, we can
produce traces that span the whole range of the complexity spectrum,
from completely predictable to completely unstructured---which makes
this an ideal testbed for this study\footnote{Predicting the
  \emph{state} of a computer, of course, would amount to solving the
  halting problem.  What we are doing here is predicting computer
  \emph{performance}, which does not violate the Rice-Shapiro
  theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

### 2014-07-16 model.tex
revisions after Trident

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To sidestep this issue, we
build forecasts in a stepwise fashion: i.e., fit the \arima model
to the existing data, use that model to perform a one-step prediction,
rebuild it using the latest observations, and iterate until the
desired prediction horizon is reached.  (For consistency, we take the
same approach with the other three models in this study as well, even
though doing so amounts to artificially hobbling LMA.)

>>>>

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge (to the mean, to a constant value,
or to a linear trend) after some number of predictions, depending on
model order.  To sidestep this issue, we build forecasts in a stepwise
fashion: i.e., fit the \arima model to the existing data, use that
model to perform a one-step prediction, rebuild it using the latest
observations, and iterate until the desired prediction horizon is
reached.  For consistency, we take the same approach with the other
three models in this study as well, even though doing so amounts to
artificially hobbling LMA.

### 2014-07-16 results.tex
revisions after Trident

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the best prediction of each time series.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  [[Say what the colored region is and why we
    included it.]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each trace from
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this relationship between WPE
and prediction error\footnote{The logarithmic fit is appropriate here
  because any signal with zero entropy should be perfectly
  predictable, and so should ideally have a MASE value of zero.}.
This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an\emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
    shaded regions?  need to check this throughout the text and in all
    captions after we finalize the plot]] is the same as in the
previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The distribution of the cluster \emph{positions} is quite different in
the three plots.  The mean MASE scores of predictions generated with
the nonlinear LMA method are generally closer to the [[dashed curve]];
the \arima method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the [[dashed curve]].  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and
deterministic\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or one that uses a
linear model (\arima)---the MASE score is much better: $0.050 \pm
0.001$.  This prediction is 20 times more accurate than a random-walk
forecast, which is more in line with the level of predictive structure
that the low WPE value suggests is present in the signal.

The [[dashed curve]] in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below [[that curve]] (as in the top and middle
images in Figure~\ref{fig:wpe_vs_mase_all}), one should try a
different method.  Conversely, the position of the LMA cluster---well
to the left of the other two, near the [[dashed curve]]---reflects
this method's ability to capture and exploit the structure that is
present in this signal.  [[Liz to wordsmith: Again, this curve was
    determined from a finite set of methods and data.  But we worked
    hard to make these experiments representative and comprehensive:
    forecast methods from simple to sophisticated; time-series data
    from a system whose behavior spans the dynamical behavior space.
    While we are cautiously opmistic about its generality, more
    exploration will be required to firm up that conclusion.  Say that
    Henon, Lorenz and SFI A fall within the colored region too, and
    that we're working on a broader study.]]

>>>>

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the best prediction of each time series.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  The shaded region boundaries reflect the
standard deviation of the model in its parameter space.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each trace from
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this relationship between WPE
and prediction error quite nicely.  The logarithmic shape of this
curve makes sense here because any signal with zero entropy should be
perfectly predictable, and so should ideally have a MASE value of
zero.  The shaded region shows the standard deviation of the model in
its parameter space: i.e., $y = a \log(b x + c)$ with $\pm$ one
standard deviation on each of the three parameters.

This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an\emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve and
shaded region is the same as in the previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The distribution of the cluster \emph{positions} is quite different in
the three plots.  The mean MASE scores of predictions generated with
the nonlinear LMA method are generally closer to the dashed curve; the
\arima method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the dashed curve.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and
deterministic\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or one that uses a
linear model (\arima)---the MASE score is much better: $0.050 \pm
0.001$.  This prediction is 20 times more accurate than a random-walk
forecast, which is more in line with the level of predictive structure
that the low WPE value suggests is present in the signal.

The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that curve (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed curve---reflects this method's
ability to capture and exploit the structure that is present in this
signal.  [[Liz to wordsmith: Again, this curve was determined from a
    finite set of methods and data.  But we worked hard to make these
    experiments representative and comprehensive: forecast methods
    from simple to sophisticated; time-series data from a system whose
    behavior spans the dynamical behavior space.  While we are
    cautiously opmistic about its generality, more exploration will be
    required to firm up that conclusion.  Say that Henon, Lorenz and
    SFI A fall within the colored region too, as do nonlinear
    transformations of \svdtwo and \svdsix.  Add here how much the fit
    changed with those additions.  As we add examples to this list,
    this empirical result will firm up.]]
